### 2018-02-20 submission/manuscript.Rmd
update abstract and intro

**Background.** An increasing body of research has found that various methodological steps along the 16S rRNA gene survey pipeline can change the observed microbial community. The most common studied methodological components include preservation media, extraction kit, bead beating time, and primers. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias. Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.

**Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR approach. First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated. Finally, the chimera prevalence correlation with number of OTUs was assessed.

@@

**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have artificially inflated values that are not comparable across studies.

@@

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results. This current study specifically addresses the PCR contribution of bias to this overall workflow.

Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases. The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017]. However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. Although these are important sources of bias they are not the only sources that should be critically examined. The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline, regardless of DNA polymerase used [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the V4 region of the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

>>>>

**Background.** Research has identified various methodological steps along the 16S rRNA gene survey pipeline that can change microbial community results. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias. Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream bacterial community results.

**Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach. First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated. Finally, the chimera prevalence correlation with number of OTUs was assessed.

@@

**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have values that are not comparable across studies.

@@

Recently, there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. Investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant focal point of this push towards standardization [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Due to this focus, 16S rRNA gene sequencing has been much maligned for introduced biases. Yet, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. For these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

The most commonly studied biases are preservation media, storage conditions, and DNA extraction kits. Although DNA extraction kits have consistently been shown to add bias to downstream analysis the current literature on preservation media and storage conditions has been mixed, with some studies showing biases while others do not [@contamination_Salter2014; @metagenomcis_bias_Costea_2017; @preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. The study of these specific biases has become so large, aggregating them all together has become a difficult task with some researchers providing resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). It is evident that these three sources of bias have been extensively covered within the literature but they are not the only sources that should be critically examined. The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-02-20 submission/manuscript.tex
update abstract and intro

\textbf{Background.} An increasing body of research has found that
various methodological steps along the 16S rRNA gene survey pipeline can
change the observed microbial community. The most common studied
methodological components include preservation media, extraction kit,
bead beating time, and primers. Although cycle number and high fidelity
(HiFi) DNA polymerase are studied less often, they are still important
sources of bias. Here, we critically examine both cycle number and HiFi
DNA polymerase for biases that may influence downstream diversity
measures of 16S rRNA gene surveys.

\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
approach. First, the number of OTUs (Operational Taxonomic Units) were
examined for both fecal samples and mock communities. Next, Bray-Curtis
index, the error rate, sequence error prevalence, and chimera prevalence
were investigated. Finally, the chimera prevalence correlation with
number of OTUs was assessed.

@@

\textbf{Conclusions.} Due to HiFi DNA polymerase dependent differences
in the number of OTUs and chimera prevalence, common diversity metrics
could have artificially inflated values that are not comparable across
studies.

@@

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced biases, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include preservation media, storage conditions, DNA
extraction kit, PCR, and sequence library preparation. Thus, for these
overlapping considerations, biases identified for 16S rRNA gene
sequencing will also likely influence metagenomic sequencing results.
This current study specifically addresses the PCR contribution of bias
to this overall workflow.

Currently, preservation media used, storage conditions, and DNA
extraction kits chosen are the most commonly studied biases. The study
of these specific biases has become so large, aggregating them all
together has become a difficult task and some researchers provide
resources to actively track new findings (e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
Within the literature, DNA extraction kits have consistently been shown
to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
2017). However, the current literature on preservation media and storage
conditions has been more mixed, with some studies showing biases while
others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
Although these are important sources of bias they are not the only
sources that should be critically examined. The type of DNA polymerase
chosen could also have a wide ranging affect on downstream results due
to error rates and chimeras that may not be easily resolved using
bioinformatic approaches.

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that the number of Operational Taxonomic Units (OTUs)
generated were not easily removed using the authors chosen bioinformatic
pipeline, regardless of DNA polymerase used (Gohl et al., 2016).
Although it is probably not surprising that normal DNA polymerase
performed worse than HiFi DNA polymerase, it is natural to extend this
line of inquiry and ask whether different HiFi DNA polymerase contribute
different biases to downstream sequencing results. There is some reason
to think that this may be the case since many of these HiFi DNA
polymerase come from different families (e.g. \emph{Taq} belongs to the
family A polymerases) and may intrinsically have different error rates
that cannot be completely removed with modifications (Ishino \& Ishino,
2014). In this study we critically examine if any of five different HiFi
DNA polymerases introduce significant biases into 16S rRNA gene surveys,
if this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.

We amplified the V4 region of the 16S rRNA gene in both fecal and mock
community samples using either Accuprime, Kappa, Q5, Phusion, or
Platinum HiFi DNA polymerase. First, we tested if we could identify
differences in the number of OTUs between the different HiFi DNA
polymerase for both the fecal and mock samples. Next, since there were
differences based on HiFi DNA polymerase, we examined if a community
measure such as the Bray-Curtis index would be affected. After assessing
these community measures we then examined if the different HiFi DNA
polymerase had different per base error rates as well as chimera
prevalence. Finally, since differences in chimera prevalence, based on
HiFi DNA polymerase, could not be completely removed using bioinformatic
approaches we tested whether the chimera prevalence correlated well with
total number of observed OTUs.

>>>>

\textbf{Background.} Research has identified various methodological
steps along the 16S rRNA gene survey pipeline that can change microbial
community results. Although cycle number and high fidelity (HiFi) DNA
polymerase are studied less often, they are still important sources of
bias. Here, we critically examine both cycle number and HiFi DNA
polymerase for biases that may influence downstream bacterial community
results.

\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same
approach. First, the number of OTUs (Operational Taxonomic Units) were
examined for both fecal samples and mock communities. Next, Bray-Curtis
index, the error rate, sequence error prevalence, and chimera prevalence
were investigated. Finally, the chimera prevalence correlation with
number of OTUs was assessed.

@@

\textbf{Conclusions.} Due to HiFi DNA polymerase dependent differences
in the number of OTUs and chimera prevalence, common diversity metrics
could have values that are not comparable across studies.

@@

Recently, there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). Investigating ways that 16S rRNA gene
surveys can be made more reproducible has been a predominant focal point
of this push towards standardization (Lauber et al., 2010; Salter et
al., 2014; Song et al., 2016; Gohl et al., 2016). Due to this focus, 16S
rRNA gene sequencing has been much maligned for introduced biases. Yet,
many of these same considerations also affect metagenomic sequencing
(Nayfach \& Pollard, 2016; Costea et al., 2017). Between the two
approaches similar bias considerations include preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. For these overlapping considerations, biases identified for
16S rRNA gene sequencing will also likely influence metagenomic
sequencing results.

The most commonly studied biases are preservation media, storage
conditions, and DNA extraction kits. Although DNA extraction kits have
consistently been shown to add bias to downstream analysis the current
literature on preservation media and storage conditions has been mixed,
with some studies showing biases while others do not (Lauber et al.,
2010; Dominianni et al., 2014; Salter et al., 2014; Sinha et al., 2015;
Song et al., 2016; Luo et al., 2016; Bassis et al., 2017; Costea et al.,
2017). The study of these specific biases has become so large,
aggregating them all together has become a difficult task with some
researchers providing resources to actively track new findings
(e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
It is evident that these three sources of bias have been extensively
covered within the literature but they are not the only sources that
should be critically examined. The type of DNA polymerase chosen could
also have a wide ranging affect on downstream results due to error rates
and chimeras that may not be easily resolved using bioinformatic
approaches.

A recent study in \emph{Nature Biotechnology} found clear differences
between normal and high fidelity (HiFi) DNA polymerase and that you
could reduce error and chimera generation by optimizing the PCR protocol
(Gohl et al., 2016). This study also found that, regardless of DNA
polymerase, the number of Operational Taxonomic Units (OTUs) generated
were not easily removed using the authors chosen bioinformatic pipeline
(Gohl et al., 2016). It is natural to extend this line of inquiry and
ask whether different HiFi DNA polymerase contribute different biases to
downstream sequencing results. There is some reason to think that this
may be the case since many of these HiFi DNA polymerase come from
different families (e.g. \emph{Taq} belongs to the family A polymerases)
and may intrinsically have different error rates that cannot be
completely removed with modifications (Ishino \& Ishino, 2014). In this
study we critically examine if any of five different HiFi DNA
polymerases introduce significant biases into 16S rRNA gene surveys, if
this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.
